Indirect Conflicts: An Exploration and Discussion of Tools, Process, and Developer Insight

### 2013-10-03 iced-2013/paper.tex
ICED 2nd draft complete.

\subsection{Grounded Theory}
We approached our study based on grounded theory as described by Corbin and Strauss~\cite{Corbin:1998:SP}.
Grounded theory is a qualitative research methodology that utilizes \textit{theoretical sampling} and
\textit{open coding} to create a theory ``grounded'' in the empirical data. For an exploratory study such as
the ours, grounded theory is well suited because it involves starting from very broad and abstract type questions and
making refinements along the way as the study progresses and hypotheses begin to take shape. Grounded theory involves
realigning the sampling criteria throughout the course of the study to ensure that participants are able to answer new
questions that have been formulated in regards to forming hypotheses. In our study being presented, data collected from
both interviews and surveys (when open ended questions were involved) was analyzed using open coding. Open coding involves
assigning codes to what participants said at a low sentence level or abstractly at a paragraph or full answer level. These
codes were defined as the study progressed and different hypotheses began to grow. We finally use axial coding it order to
link our defined codes through categories of grounded theory such as context and consequences. The ultimate goal of grounded 
theory is to produce a theory that is tied to the empirical data collected; our final theories can be seen in
findings~\ref{itm:f1} -~\ref{itm:f9}. In Section~\ref{sec:eval}, we give a brief evaluation of our studying using 
3 criteria that are commonly used in evaluating grounded theory studies.

@@

From the interviewed participants, 63\% of developers believe that a large contributing factor to the cause
of indirect conflicts comes from the changing of a software object's contract. Object contracts are, in a sense,
what a software object guarantees, meaning how the input, output, or how any aspect of the object is guaranteed
to work; made famous by Eiffle Software's~\footnote{http://www.eiffel.com/} ``Design by Contract''\texttrademark. 
In light of object contracts, 73\% of interviewed developers gave examples of indirect conflicts they had experienced
which stemmed from not understanding the far reaching ramifications of a change being made to an object contract
towards the rest of the project. Of those 73\%, examples were broken down as follows. 21\% of examples dealt
with the changing of legacy code, with one developer saying ``legacy code does not change because developers 
are afraid of the long range issues that may arise from that change''.

From these listed results, we see a common theme of awareness is obvious. Developers cause indirect conflicts because they do not
fully understand the ramifications of their source code change, especially when a software object's contract is being changed. 
This was largely apparent when developers were dealing with legacy code,
or another developer's code in general. The non-understanding of how implementation decisions were made is a large factor and that changing
older code was quite prone to indirect conflicts as they were unsure of where the code was being used. This presents an immediate problem regarding
indirect conflicts in software maintenance and we would think the simple fix would be static analysis tools for source
code to allow developers to identify which parts of the project they are affecting with a change. However, as we found out this is not always
the easy fix. For starters, not all programming languages are subject to strong static analysis if any, while some projects involve many
cross language dependencies which also breaks down any potential for static analysis on these software dependencies. Another potential solution
to this issue could be formal documentation such as UML. However, from Petre~\cite{Petre:2013:UP} we see that UML is not used so much in practice.
These short comings leave the door open for other type of pro-developer documentation or tools to allow quick understanding of code changes and
how they may impact cross team, cross project, or other dependencies. At any rate, it is clear that developers do not posses the tools needed
to quickly evaluate a software change (especially when contracts change) from a wide spectrum of dependencies that may not have the ability to
be statically analyzed.

@@

From the interviewed participants, 58\% of developers explained that indirect conflicts occur ``all the time'' in
their development life cycle with a minimum occurrence of once a week. Of those 58\%, developers noted that a
lot of their development time is put into understanding code to avoid indirect conflicts, with more serious issues tending
to occur once a month, and that the conflicts that do occur tend to be quite unique from each other. Right away we can see a serious issue with the standard
development work flow. Developers are not actually spending their time developing, rather they are spending large amounts of time trying to avoid indirect
conflicts. This shows that indirect conflicts are a serious threat to developer productivity and are a problem worth study and investigation.

@@

From the interviews and the surveys combined we can make the deduction that indirect conflicts are more likely to occur early 
in a projects development cycle as opposed to later. This must be a major concern for any indirect conflict tools moving forward. Since many
of the tools listen in Section~\ref{sec:intro} implement some sort of filtering system or taxonomy to avoid information overload, this
finding should be included in future workings. Knowing that indirect conflicts occur more frequently early in a development cycle will let
these tools know that those reported later in the development life cycle have a high probability (compared to without this knowledge) of
being a true positive. This is a major concern when dealing with information overload and to provide developers with what they may find
relevant to their work. The results in in Table~\ref{tab:often} also suggest that preemptive indirect conflict tools may be more accurate
towards the mature point of a project, but have potentially more use earlier where the indirect conflicts are more likely to occur.
While what the ``mature point'' of a project is could be speculated upon, we suggest it is caused for one of two reasons, The
first would be that the project's source code has stabilized quite a bit in terms of large changes not occurring as much. This could be a result
of a project being released to a customer or because it is now being used as part of many other projects. The second would be that a change in
development process has occurred in that source code is not being changed so much as newer code is being added or old code is being deleted. This
would cause a slow in indirect conflicts as developers interviewed said that changes to code, not the addition of or deletion of code is prone
to indirect conflicts. In the end, developers survey pointed out that before the ``mature
point'' indirect conflicts are like to occur daily and weekly (32\% and 18\%) as opposed to after the ``mature point'' when developers
said they are more likely to occur on a monthly or even yearly frequency (18\% and 12\%).

Of those surveyed that could not give direct answers to the questions in Table~\ref{tab:often}, the more common answers were 
as follows: ``People rushing to meet [the release date] is where most of my indirect conflicts occur.'', ``Indirect conflicts
after a release depends on how well the project was built at first'', ``They tend to slow down a bit after a major release, 
unless the next release is a major rework.'', and ``Conflicts continue to occur at roughly the same rate after the first release,
with spikes during large revamps or the implementation of cross-cutting new features.''. These results can be argued to show that the occurance
of indirect conflicts can largely depend on the management, and deadlines of projects. These are important factors to consider when trying
to define a ``mature point''. A large refactoring or rewrite can reset the ``mature point'' of a project and should be considered for
any taxonomies created for preemptive indirect conflict tools.

@@

From Table~\ref{tab:env}, is can clearly be seen that as a project becomes larger and more and more developers are added, even to the point
that multiple teams have to be formed, indirect conflicts become more likely to occur. However, this is not to say that indirect conflicts
do not occur at the lower number of developer level. Even when developing individual projects, 43\% of developers agreed or strongly agreed
that indirect conflicts are likely to occur. Once more than one developer is associated with a project, or a project has multiple teams, 
67\% and 74\% of developers agreed or strongly agreed that indirect conflicts are likely to occur. These results show that indirect conflicts
seem to be compounded by a multiple human factor found in parallel development. The lack of awareness as to other developers work in a project 
again comes to light. We would think that developers would leverage the other developer's knowledge in their projects to avoid indirect conflicts
but as we will see most developers tend not leverage their social connections within a project to help solve indirect conflicts. This could be
a potential issue of social structure in that developers simply do not have the means or will to talk to their colleagues, or it could be that
developers heavily leverage the notion of software stigmergy as shown by Bolici et. al~\cite{Bolici}. Creating effective communication around
technical dependencies between developers and teams through either explicit or implicit means appears to be a large task to accomplish but a
carsickness task as shown by these result being that static analysis breaks down at these levels.

@@

Lastly, pure developer experience was mentioned
as a source to prevent indirect conflicts. 37\% of developers mentioned that when planning code changes, either a very experienced member
of the project was involved in the planning and has duties to foresee any indirect conflicts that may arise, or that while implementing
a code change developers must use their personal knowledge to predict where indirect conflicts will occur. While most interviewed developers
had such an expert in their software projects, they admitted to not being able to foresee all indirect conflicts that may occur, especially
when talking about external teams or projects who may or may not use their final product. This raises the question of how much planning
by expert developers should really go into avoiding indirect conflicts? Dromey~\cite{} raises the debate of prevention versus cure of software
defects and how difficult a problem it is to measure. Should developers be using certain tools to plan their changes or should they be even
planning at all? Which method provides the best software quality for the minimum effort. These are not easy questions to answer but may
be key in improving developer productivity when concerning indirect conflicts.

@@

To ensure the correctness of the results, we also linked all findings in Section~\ref{sec:disc} to either a majority of agreeing
responses on a topic or to a large amount of direct quotes presented by participants.

@@

Hattori et. al.~\cite{Hattori:2012:ICG} found, through qualitative user studies, that developers tend to use the bare amount of communication
towards other developers in order to solve direct conflicts, and take the same approach of only communicating once a conflict has arose. This
is the same mentality found in this paper as ``I work until something breaks''. Hattori et. al. also show that with direct conflicts, the
sooner preemptive information if available to developers the more they will communicate and either avoid or easily solve their code merges.
This is the situation many indirect conflict tools have strived for. In terms of communication, Bolici et. al~\cite{Bolici} give the 
possibility of developer stigmergy as the reason that developers do not need to explicitly communicate with each other about issues. The idea
that developers use artifacts left behind by other developers to solve their issues could play a large role in a lack of formal communication.

>>>>

\subsection{Grounded Theory}
We approached our study based on grounded theory as described by Corbin and Strauss~\cite{Corbin:1998:SP}.
Grounded theory is a qualitative research methodology that utilizes \textit{theoretical sampling} and
\textit{open coding} to create a theory ``grounded'' in the empirical data. For an exploratory study such as
the ours, grounded theory is well suited because it involves starting from very broad and abstract type questions and
making refinements along the way as the study progresses and hypotheses begin to take shape. Grounded theory involves
realigning the sampling criteria throughout the course of the study to ensure that participants are able to answer new
questions that have been formulated in regards to forming hypotheses. In our study being presented, data collected from
both interviews and surveys (when open ended questions were involved) was analyzed using open coding. Open coding involves
assigning codes to what participants said at a low sentence level or abstractly at a paragraph or full answer level. These
codes were defined as the study progressed and different hypotheses began to grow. We finally use axial coding it order to
link our defined codes through categories of grounded theory such as context and consequences. The ultimate goal of grounded 
theory is to produce a theory that is tied to the empirical data collected; our final theories can be seen in
findings~\ref{itm:f1} -~\ref{itm:f8}. In Section~\ref{sec:eval}, we give a brief evaluation of our studying using 
3 criteria that are commonly used in evaluating grounded theory studies.

@@

From the interviewed participants, 63\% of developers believe that a large contributing factor to the cause
of indirect conflicts comes from the changing of a software object's contract. Object contracts are, in a sense,
what a software object guarantees, meaning how the input, output, or how any aspect of the object is guaranteed
to work; made famous by Eiffle Software's~\footnote{http://www.eiffel.com/} ``Design by Contract''\texttrademark. 
In light of object contracts, 73\% of interviewed developers gave examples of indirect conflicts they had experienced
which stemmed from not understanding the far reaching ramifications of a change being made to an object contract
towards the rest of the project. Of those 73\%, 21\% dealt
with the changing of legacy code, with one developer saying ``legacy code does not change because developers 
are afraid of the long range issues that may arise from that change''.

From these listed results, we see a common theme of awareness is obvious. Developers cause indirect conflicts because they do not
fully understand the ramifications of their source code change, especially when a software object's contract is being changed. 
This was largely apparent when developers were dealing with legacy code,
or another developer's code in general. The non-understanding of how implementation decisions were previously made is a large factor and that changing
older code was quite prone to indirect conflicts as they were unsure of where the code was being used. This presents an immediate problem regarding
indirect conflicts in software maintenance and we would think the simple fix would be static analysis tools for source
code to allow developers to identify which parts of the project they are affecting with a change. However, as we found out this is not always
the easy fix. Not all programming languages are subject to strong static analysis if any, while some projects involve many
cross language dependencies which also breaks down any potential for static analysis on these software dependencies. Another potential solution
to this issue could be formal documentation such as UML. However, from Petre~\cite{Petre:2013:UP} we see that UML is not used so much in practice.
These short comings leave the door open for other type of pro-developer documentation or tools to allow quick understanding of code changes and
how they may impact cross team, cross project, or other dependencies. At any rate, it is clear that developers do not posses the tools needed
to quickly evaluate a software change (especially when contracts change) from a wide spectrum of dependencies that may not have the ability to
be analyzed by traditional means. This finding lays the foundation for the reality of indirect conflicts occurring in industry,

@@

From the interviewed participants, 58\% of developers explained that indirect conflicts occur ``all the time'' in
their development life cycle with a minimum occurrence of once a week. Of those 58\%, developers noted that a
lot of their development time is put into understanding code to avoid indirect conflicts, with more serious issues tending
to occur once a month, and that the conflicts that do occur tend to be quite unique from each other. Right away we can see a serious issue with the standard
development work flow. Developers are not actually spending their time developing, rather they are spending large amounts of time trying to avoid indirect
conflicts. This shows that indirect conflicts are a serious threat to developer productivity and are a problem worth study and investigation. This also
raises the point of how developers should be spending their time: preventing, or solving indirect conflicts.

@@

From the interviews and the surveys combined we can make the deduction that indirect conflicts are more likely to occur early 
in a projects development cycle as opposed to later. This must be a major concern for any indirect conflict tools moving forward. Since many
of the tools listen in Section~\ref{sec:intro} implement some sort of filtering system or taxonomy to avoid information overload, this
finding should be included in future workings. Knowing that indirect conflicts occur more frequently early in a development cycle will let
these tools know that those reported later in the development life cycle have a higher probability (compared to without this knowledge) of
being a true positive. This is a major concern when dealing with information overload and to provide developers with what they may find
relevant to their work. The results in in Table~\ref{tab:often} also suggest that preemptive indirect conflict tools may be more accurate
towards the mature point of a project, but have potentially more use earlier where the indirect conflicts are more likely to occur.
While what the ``mature point'' of a project is could be speculated upon, we suggest it is caused for one of two reasons, The
first would be that the project's source code has stabilized quite a bit in terms of large changes not occurring as much. This could be a result
of a project being released to a customer or because it is now being used as part of many other projects. The second would be that a change in
development process has occurred in that source code is not being changed so much as newer code is being added or old code is being deleted. This
would cause a slow in indirect conflicts as developers interviewed said that changes to code, not the addition of or deletion of code is prone
to indirect conflicts. In the end, developers survey pointed out that before the ``mature
point'' indirect conflicts are like to occur daily and weekly (32\% and 18\%) as opposed to after the ``mature point'' when developers
said they are more likely to occur on a monthly or even yearly frequency (18\% and 12\%).

Of those surveyed that could not give direct answers to the questions in Table~\ref{tab:often}, the more common answers were 
as follows: ``People rushing to meet [the release date] is where most of my indirect conflicts occur.'', ``Indirect conflicts
after a release depends on how well the project was built at first'', ``They tend to slow down a bit after a major release, 
unless the next release is a major rework.'', and ``Conflicts continue to occur at roughly the same rate after the first release,
with spikes during large revamps or the implementation of cross-cutting new features.''. These results can be argued to show that the occurrence
of indirect conflicts can largely depend on the management, and deadlines of projects. These are important factors to consider when trying
to define a ``mature point''. A large refactoring or rewrite can reset the ``mature point'' of a project and should be considered for
any taxonomies created for preemptive indirect conflict tools. The ``mature point'' of a project (or multiple mature points) should be
considered when building a taxonomy for indirect conflict tools.

@@

From Table~\ref{tab:env}, is can clearly be seen that as a project becomes larger and more and more developers are added, even to the point
that multiple teams have to be formed, indirect conflicts become more likely to occur. However, this is not to say that indirect conflicts
do not occur at the lower number of developer level. Even when developing individual projects, 43\% of developers agreed or strongly agreed
that indirect conflicts are likely to occur. Once more than one developer is associated with a project, or a project has multiple teams, 
67\% and 74\% of developers agreed or strongly agreed that indirect conflicts are likely to occur. These results show that indirect conflicts
seem to be compounded by a multiple human factor found in parallel development. The lack of awareness as to other developers work in a project 
again comes to light. We would think that developers would leverage the other developer's knowledge in their projects to avoid indirect conflicts
but as we will see most developers tend not leverage their social connections within a project to help solve indirect conflicts. This could be
a potential issue of social structure in that developers simply do not have the means or will to talk to their colleagues, or it could be that
developers heavily leverage the notion of software stigmergy as shown by Bolici et. al~\cite{Bolici:2009}. Creating effective communication around
technical dependencies between developers and teams through either explicit or implicit means appears to be a large task to accomplish but a
carsickness task as shown by these result being that static analysis breaks down at these levels.

This finding characterizes that indirect conflicts are a frequent problem in industry. It also shows what environments are more prone to indirect
conflicts and should be handled with more delicate care than others.

@@

Lastly, pure developer experience was mentioned
as a source to prevent indirect conflicts. 37\% of developers mentioned that when planning code changes, either a very experienced member
of the project was involved in the planning and has duties to foresee any indirect conflicts that may arise, or that while implementing
a code change developers must use their personal knowledge to predict where indirect conflicts will occur. While most interviewed developers
had such an expert in their software projects, they admitted to not being able to foresee all indirect conflicts that may occur, especially
when talking about external teams or projects who may or may not use their final product. This raises the question of how much planning
by expert developers should really go into avoiding indirect conflicts? Dromey~\cite{Dromey:2003} raises the debate of prevention versus cure of software
defects and how difficult a problem it is to measure. Should developers be using certain tools to plan their changes or should they be even
planning at all? Which method provides the best software quality for the minimum effort. These are not easy questions to answer but may
be key in improving developer productivity when concerning indirect conflicts.

@@

To ensure the correctness of the results, we also linked all findings in Section~\ref{sec:results} to either a majority of agreeing
responses on a topic or to a large amount of direct quotes presented by participants.

@@

Hattori et. al.~\cite{Hattori:2012:ICG} found, through qualitative user studies, that developers tend to use the bare amount of communication
towards other developers in order to solve direct conflicts, and take the same approach of only communicating once a conflict has arose. This
is the same mentality found in this paper as ``I work until something breaks''. Hattori et. al. also show that with direct conflicts, the
sooner preemptive information if available to developers the more they will communicate and either avoid or easily solve their code merges.
This is the situation many indirect conflict tools have strived for. In terms of communication, Bolici et. al~\cite{Bolici:2009} give the 
possibility of developer stigmergy as the reason that developers do not need to explicitly communicate with each other about issues. The idea
that developers use artifacts left behind by other developers to solve their issues could play a large role in a lack of formal communication.
